---
title: "Data Cleaning: FiveThirtyEight"
output: 
  prettydoc::html_pretty:
    theme: leonids
    css: styles.css
---

[Go Back](github_projects.html)

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, message = FALSE)
```

## Overview

This project is part of the bigger project of which I thoroughly walked through steps in data cleaning & wrangling process. I then did some descriptive explorations of the dataset. In this project, I used the pols-month.csv, unemployment.csv, and snp.csv data from [FiveThirtyEight](https://p8105.com/dataset_fivethirtyeight.html). This project utilized R version 4.3.1, and R packages `tidyverse` and `dplyr`. 

```{r github_repository, echo=FALSE}
usethis::use_git_config(
    user.name = "Xuan Lu", 
    user.email = "xuan.lu.080229@gmail.com"
  )
```

```{r library_packages, message = FALSE}
library(tidyverse)
library(dplyr)
```

## STEP 1: Clean dataset pols-month.csv

Directions: Use `separate()` to break up the variable mon into integer variables *year*, *month*, and *day*; replace month number with month name; create a *president* variable taking values *gop* and *dem*, and remove *prez_dem* and *prez_gop*; and remove the *day* variable.

```{r clean pols_month, message = FALSE}
pols_month <- 
  read_csv("./fivethirtyeight_datasets/pols-month.csv") %>% 
  separate(mon, 
         into = c("year","month_num","day"), 
         convert = TRUE) %>% 
  mutate("month" = month.name[month_num]) %>%
  mutate(president = recode(prez_gop, "0" = "dem", "1" = "gop", "2" = "gop")) %>%
  select(year, month, everything(), -day, -prez_gop, -prez_dem, -month_num) 
pols_month
```

## STEP 2: Clean dataset snp.csv

Directions: Use a similar process to the above. For consistency across datasets, arrange according to *year* and *month*, and organize so that *year* and *month* are the leading columns.

```{r clean snp, message = FALSE }
snp <- 
  read_csv("./fivethirtyeight_datasets/snp.csv") %>%
  separate(date, into = c("month", "day", "year"), sep = "/", convert = TRUE) %>%
  arrange(year, month) %>%
  mutate("month" = month.name[month]) %>%
  select(year, month, close) 
snp
```

## STEP 3: Tidy dataset unemployment.csv

Directions: Tidy the unemployment data so that it can be merged with the previous datasets. This process will involve switching from `“wide”` to `“long”` format; ensuring that key variables have the same name; and ensuring that key variables take the same values.

```{r clean unemployment, message = FALSE}
unemployment <- 
  read_csv("./fivethirtyeight_datasets/unemployment.csv") %>%
  pivot_longer(Jan:Dec, names_to = "month_abb", values_to = "unemployment") %>%
  mutate("month" = factor(month_abb, levels = month.abb, labels = month.name)) %>%
  rename(year = Year) %>%
  select(year, month, unemployment)
unemployment
```

## STEP 4: Merge datasets

Directions: Join the datasets by merging *snp* into *pols*, and merging *unemployment* into the result.

```{r merging, message = FALSE }
data_q1 <- full_join(pols_month, snp, by = c("year","month")) %>%
  full_join(., unemployment, by = c("year","month"))
summary(data_q1)
```

**pols_month**
There are `r nrow(pols_month)` observations and `r ncol(pols_month)` variables in the cleaned *pols_month* dataset, ranges between years of `r min(pols_month$year)` and `r max(pols_month$year)`. Variables are: *`r colnames(pols_month)`*.

**snp**
There are `r nrow(snp)` observations and `r ncol(snp)` variables in the cleaned *snp* dataset, ranges between years of `r min(snp$year)` and `r max(snp$year)`. Variables are: *`r colnames(snp)`*.

**unemployment**
There are `r nrow(unemployment)` observations and `r ncol(unemployment)` variables in the cleaned *unemployment* dataset, ranges between years of `r min(unemployment$year)` and `r max(unemployment$year)`. Variables are: *`r colnames(unemployment)`*.

**merged dataset `data_q1`**
There are a total of `r nrow(data_q1)` observations and `r ncol(data_q1)` variables in the final merged dataset, ranges between years of `r min(data_q1$year)` and `r max(data_q1$year)`. Variables are: *`r colnames(data_q1)`*. Note that I used the `full_join` function with the combination of *year* and *month* as unique identifiers here. This way, all data entries from all tables will be included into the final merged dataset.

## STEP 5: Save cleaned file

We can save our cleaned dataset as follows: 

```{r export final result data_q1 into a CSV file to my data directory,  message = FALSE}
write.csv(data_q1, file = "data_q1.csv")
```