---
title: "APOE4, amyloid β 42/40 ratio, and Alzheimer’s disease"
output: 
  prettydoc::html_pretty:
    theme: leonids
    css: styles.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

This project is part of the bigger project of which I thoroughly walked through steps in data cleaning & wrangling process. I then did some descriptive explorations of the dataset. In this project, I used data collected in an observational study to understand the trajectory of Alzheimer’s disease (AD) biomarkers, retried from [HERE](https://p8105.com/data/data_mci.zip). This project utilized R version 4.3.1, and R packages `tidyverse` and `dplyr`. 

```{r library_packages, message = FALSE}
library(tidyverse)
library(dplyr)
```

## STEP 1: Clean Baseline Dataset

Directions: Import, clean, and tidy the dataset of baseline demographics. Ensure that sex and APOE4 carrier status are appropriate encoded (i.e. not numeric), and remove any participants who do not meet the stated inclusion criteria (i.e. no MCI at baseline). 

```{r import and clean the dataset of baseline demographics, message = FALSE, warning = FALSE}
mci_baseline_description <- read_csv("./data_mci/MCI_baseline.csv") %>% colnames()
mci_baseline_description #checking variable descriptions

mci_baseline <- 
  #use second row as header
  read_csv("./data_mci/MCI_baseline.csv", skip = 1) %>% 
  #clean variable names to snake_case
  janitor::clean_names() %>% 
  #recode sex & apoe4 as character variables
  mutate(sex = recode(sex, "1" = "Male", "0" = "Female")) %>% 
  mutate(apoe4 = recode(apoe4, "1" = "APOE4 carrier", "0" = "APOE4 non-carrier")) %>%
  filter(age_at_onset > current_age | age_at_onset == ".") %>%
  mutate(age_at_onset = as.numeric(age_at_onset))
summary(mci_baseline)
```

When importing the dataset, we can notice that the default header of the uncleaned dataset are actually descriptions of each variable. Printing the dataset out, we would notice that the actual header is actually the second row: *`r colnames(mci_baseline)`*. 

A total of `r nrow(mci_baseline)` participants were recruited, and of these participants, `r mci_baseline %>% 
filter(!is.na(age_at_onset)) %>% 
nrow()` participants developed MCI. The average baseline age is `r mean(pull(mci_baseline, current_age))`, with a range of `r range(pull(mci_baseline, current_age))[1]` to `r range(pull(mci_baseline, current_age))[2]`. `r (count(filter(mci_baseline, apoe4 == "APOE4 carrier" & sex == "Female"))/count(filter(mci_baseline, sex == "Female")))*100`% of women are APOE4 carriers.

## STEP 2: Clean Amyloid Dataset

Directions: Import, clean, and tidy the dataset of longitudinally observed biomarker values; comment on the steps on the import process and the features of the dataset.

```{r import and clean the dataset of longitudinally observed biomarker values, message = FALSE, warning = FALSE}
mci_amyloid_description <- read_csv("./data_mci/mci_amyloid.csv") %>% colnames()
mci_amyloid_description #checking variable descriptions

mci_amyloid <- 
  #use second row as header
  read_csv("./data_mci/mci_amyloid.csv", skip = 1) %>% 
  #clean variable names to snake_case
  janitor::clean_names() %>%  
  #rename variables to match baseline dataset
  rename(id = study_id, time_0 = baseline) %>% 
  #change character variables of amyroid ratios to numeric
  mutate(time_0 = as.numeric(time_0), time_2 = as.numeric(time_2), time_4 = as.numeric(time_4),
         time_6 = as.numeric(time_6), time_8 = as.numeric(time_8)) %>% 
  #tidy dataset
  pivot_longer(., time_0:time_8, names_to = "years_elapsed_since_baseline", values_to = "amyroid_ratio", names_prefix = "time_")  
summary(mci_amyloid)
```

There are a total of `r nrow(mci_amyloid)` observations, `r length(unique(mci_amyloid$id))` unique participants and `r ncol(mci_amyloid)` variables in this dataset. When importing the dataset, we can notice that the default header of the uncleaned dataset are actually descriptions of each variable. Printing the dataset out, we would notice that the actual header is actually the second row: *`r colnames(mci_amyloid)`*. To tidy the dataset, I used `pivot_longer` function to change several columns of different times of followup to rows. 

There are `r length(unique(subset(mci_baseline, !id %in% mci_amyloid$id)$id))` participants who appeared in only the baseline dataset but not the amyloid dataset, and `r length(unique(subset(mci_amyloid, !id %in% mci_baseline$id)$id))` participants who appeared in only the amyloid dataset not the baseline dataset. To ensure that only participants with entries in both dataset are included into the final dataset, I will merge the two datasets using the `inner_join` function. 

## Step 3: Merge the two datasets

Before merging, it is always good to use `anti_join` to check if there are entries in one dataset that do not have a corresponding match in the other based on specified key columns. This is especially important for datasets with matching unique IDs. 

```{r}
not_matched <- anti_join(mci_baseline, mci_amyloid)
not_matched
```

We see that there are `r nrow(not_matched)` entries in mci_baseline that do not have a match in mci_amyloid. This hints to loss-to-follow-up and may warrant concerns regarding selection bias. Keep this noted, let's continue on merging. 

```{r merge the baseline & amyloid datasets,  message = FALSE}
data_q3 <- inner_join(mci_baseline, mci_amyloid, by = "id")
summary(data_q3)
```

There are `r nrow(data_q3)` observations, `r length(unique(data_q3$id))` unique participants, and `r ncol(data_q3)` variables in the final merged dataset for MCI. Variables include *`r colnames(data_q3)`*. 
An Aβ42/40 ratio <0.150 suggests a higher risk of having of AD pathology compared to higher values. 
2 years after baseline, `r round(length(unique(filter(data_q3, amyroid_ratio < 0.15 & years_elapsed_since_baseline == 2))$id)/length(unique(filter(data_q3, years_elapsed_since_baseline == 2))$id)*100, 2)`% of participants have An Aβ42/40 ratio <0.150. 
4 years after baseline, the proportion increased to `r round(length(unique(filter(data_q3, amyroid_ratio < 0.15 & years_elapsed_since_baseline == 4))$id)/length(unique(filter(data_q3, years_elapsed_since_baseline == 4))$id)*100, 2)`%. 
6 years after baseline, the proportion increased to `r round(length(unique(filter(data_q3, amyroid_ratio < 0.15 & years_elapsed_since_baseline == 6))$id)/length(unique(filter(data_q3, years_elapsed_since_baseline == 6))$id)*100, 2)`%. 
8 years after baseline, the proportion increased to `r round(length(unique(filter(data_q3, amyroid_ratio < 0.15 & years_elapsed_since_baseline == 8))$id)/length(unique(filter(data_q3, years_elapsed_since_baseline == 8))$id)*100, 2)`%. 
Among those that are female, the proportion of having an Aβ42/40 ratio <0.150 8 years after baseline is `r round(length(unique(filter(data_q3, amyroid_ratio < 0.15 & years_elapsed_since_baseline == 8 & sex == "Female"))$id)/length(unique(filter(data_q3, years_elapsed_since_baseline == 8 & sex == "Female"))$id)*100, 2)`%. 
This proportion is `r round(length(unique(filter(data_q3, amyroid_ratio < 0.15 & years_elapsed_since_baseline == 8 & sex == "Male"))$id)/length(unique(filter(data_q3, years_elapsed_since_baseline == 8 & sex == "Male"))$id)*100, 2)`% for males. 

## STEP 4: Save cleaned file

We can save our cleaned dataset as follows: 

```{r export final result data_q3 into a CSV file to my data directory,  message = FALSE}
write.csv(data_q3, file = "data_q3.csv")
```
