---
title: 'Data Cleaning: Trashwheel'
output: 
  prettydoc::html_pretty:
    theme: leonids
    css: styles.css
---

[Go Back](github_projects.html)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

This project is part of the bigger project of which I thoroughly walked through steps in data cleaning & wrangling process. I then did some descriptive explorations of the dataset. In this project, I used the data from [Mr. Trash Wheel](https://p8105.com/dataset_mr_trash_wheel.html). This project utilized R version 4.3.1, and R packages `tidyverse` and `dplyr`. 

```{r library_packages, message = FALSE}
library(tidyverse)
library(dplyr)
```

## STEP 1: read & clean Mr. Trash Wheel sheet

Directions: 

* specify the sheet in the Excel file and to omit non-data entries (rows with notes / figures; columns containing notes) using arguments in `read_excel`
* use reasonable variable names
* omit rows that do not include dumpster-specific data
* The data include a column for the (approximate) number of homes powered. This calculation is described in the Homes powered note, but not applied to every row in the dataset. Update the data to include a new homes_powered variable based on this calculation.

```{r read and clean the Mr. Trash Wheel sheet, message = FALSE}
mr_trash_wheel <- 
  readxl::read_excel("202309\ Trash\ Wheel\ Collection\ Data.xlsx", 
                     sheet = "Mr. Trash Wheel", col_names = TRUE, range = "A2:N586") %>%
  janitor::clean_names() %>%
  mutate("day" = format(date, format="%d")) %>%
  mutate("year" = format(year, format="%Y")) %>%
  select(dumpster, year, month, day, everything(), -date) %>%
  mutate("homes_powered" = (weight_tons*500)/30) %>%
  mutate("dumpster" = paste("mr", dumpster, sep="_"))
head(mr_trash_wheel)
```

I changed the *date* variable in this sheet to *day* as there are *year* and *month* variables. I also changed the format of *year* to character. The Mr. Trash Wheel sheet contains `r ncol(mr_trash_wheel)` variables and `r nrow(mr_trash_wheel)` observations. Variables in this sheet include: *`r colnames(mr_trash_wheel)`*. Variable *dumpster* is the **primary key** that contains values unique to each observation. 

## STEP 2: read & clean Professor Trash Wheel sheet

```{r read and clean the Professor Trash Wheel sheet, message = FALSE}
professor_trash_wheel <- 
  readxl::read_excel("202309\ Trash\ Wheel\ Collection\ Data.xlsx", 
                     sheet = "Professor Trash Wheel", col_names = TRUE, range = "A2:M108") %>%
  janitor::clean_names() %>%
  mutate("day" = format(date, format="%d")) %>%
  mutate("year" = format(year, format="%Y")) %>%
  select(dumpster, year, month, day, everything(), -date) %>%
  mutate("homes_powered" = (weight_tons*500)/30) %>%
  mutate("dumpster" = paste("professor", dumpster, sep="_"))
head(professor_trash_wheel)
```

I changed the *date* variable in this sheet to *day* as there are *year* and *month* variables. I also changed the format of *year* to character. The Professor Trash Wheel sheet contains `r ncol(professor_trash_wheel)` variables and `r nrow(professor_trash_wheel)` observations. Variables in this sheet include: *`r colnames(professor_trash_wheel)`*. Variable *dumpster* is the **primary key** that contains values unique to each observation. 

## STEP 3: read & clean Gwynnda Trash Wheel sheet

```{r read and clean the Gwynnda Trash Wheel sheet, message = FALSE}
gwynnda_trash_wheel <- 
  readxl::read_excel("202309\ Trash\ Wheel\ Collection\ Data.xlsx", 
                     sheet = "Gwynnda Trash Wheel", col_names = TRUE, range = "A2:L157") %>%
  janitor::clean_names() %>%
  mutate("day" = format(date, format="%d")) %>%
  mutate("year" = format(year, format="%Y")) %>%
  select(dumpster, year, month, day, everything(), -date) %>%
  mutate("homes_powered" = (weight_tons*500)/30) %>%
  mutate("dumpster" = paste("gwynnda",  dumpster, sep="_"))
head(gwynnda_trash_wheel)
```

I changed the *date* variable in this sheet to *day* as there are *year* and *month* variables. I also changed the format of *year* to character. The Professor Trash Wheel sheet contains `r ncol(gwynnda_trash_wheel)` variables and `r nrow(gwynnda_trash_wheel)` observations. Variables in this sheet include: *`r colnames(gwynnda_trash_wheel)`*. Variable *dumpster* is the **primary key** that contains values unique to each observation. 

## STEP 4: merge the three sheets

```{r merge the three sheets, message = FALSE}
data_q2 <-
  full_join(mr_trash_wheel, professor_trash_wheel) %>%
  full_join(., gwynnda_trash_wheel)
summary(data_q2)
```

Note that I used the `full_join` function here. This way, all data entries from all tables will be included into the final merged dataset. The merged data contains `r ncol(data_q2)` variables and `r nrow(data_q2)` observations. Variables in this sheet include: *`r colnames(data_q2)`*. Variable *dumpster* is the **primary key** that contains values unique to each observation.

```{r setting, message=FALSE, echo=FALSE, warning=FALSE}
options(scipen = 999)
```

The total weight of trash collected by **Professor Trash Wheel** is `r weight_sum <- data_q2 |> filter(grepl("professor", dumpster)) 
sum(pull(weight_sum, weight_tons))`. The total number of cigarette butts collected by **Gwynnda in July of 2021** is `r cig_sum <- data_q2 %>% filter(grepl("gwynnda", dumpster)) %>% filter(month == "July") %>% filter(year == "2021") 
sum(pull(cig_sum, cigarette_butts))`.

## STEP 5: Save cleaned file

We can save our cleaned dataset as follows: 

```{r export final result data_q2 into a CSV file to my data directory,  message = FALSE}
write.csv(data_q2, file = "data_q2.csv")
```